Data Management \\ and LSST Special Programs

### 2017-05-10 ms.tex
major revision

\documentclass[DM,lsstdraft,toc]{lsstdoc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{latexsym}
\usepackage{color}
% black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, blue, orange, pink, purple, red, teal, violet, white, yellow.

@@

\author{M.~L.~Graham}

@@

\setDocAbstract{\textbf{WORKING DRAFT } but intended to be an overview of any and all potential requirements on data management (DM) from special programs (SP) such as the deep drilling fields (DDF) or mini-surveys (MS). We also summarize the plans for how the special programs data will be integrated into the wide-fast-deep (WFD) Main Survey (Levels 1 and 2). It is the intent that this document evolve to become an internal catalog of change requests to the DM Level 1 and 2 pipelines and products, and a community resource for future special program white papers and preparations for Level 3 pipelines. \textbf{Preliminary action items are listed in Section \ref{ssec:intro_action}.}

\medskip MLG Summary: The very high level requirements for Level 3 and Special Programs are written down in the DMSR, but they're broad. The next document down-flow, the LDM-151, is focused on the Level 1/2 algorithms with no accountability to the DMSR requirements related to SP and Level 3 (and that's fine). LDM-151, Section 7, mentions the DAX and SUIT, and then also mentions that it will be detailed in a different document. At this stage it seems it is maybe a mite too early to conclude anything, but we can go through the "Action Items" and try to resolve the questions that exist, for now.
}

@@

\noindent The current status of this work is that it is an internal, ongoing study with several interim preliminary action items or outstanding questions so far (Section \ref{ssec:intro_action}).

\noindent The purpose of this study is to: \\
(1) summarize the current plans of the DM pipelines with respect to special programs, \\
(2) summarize the processing that is required to enable science for special programs, \\
(3) assess whether the DM pipelines meet the expected needs of special programs, and \\
(4) ensure any necessary changes get written into the requirements, designs, and plans.

\noindent This will be accomplished by: \\
(1) starting with the DPDD, whitepapers, etc., and compiling relevant information, \\
(2) discussing initial issues internally with DM (Mario, K-T, Tim), and then \\
(3) ensure the identified needs become specified requirements (i.e., action items Section \ref{ssec:intro_action}).

@@

\subsection{Current Action Items}\label{ssec:intro_action}

Action items for Data Management can be done internally as needed -- some are just questions of things to investigate. Items that involve gather feedback from the Science Collaborations should probably wait and happen at the same time as the next calls for community input (see Section \ref{ssec:intro_forum}).

\noindent \textbf{Data Management: } \\
\textbf{(1)} Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time than the WFD images, by user request. (Section \ref{ssec:dmdocs_review}) \\
\textbf{(2)} Confirm if the budget requires additional computational resources for Special Programs and Level 3 processing. (Section \ref{ssec:dmdocs_review}) \\
\textbf{(3)} Find out the thermal tolerances and determine the minimum and maximum exposure times (and/or cadence of short exposures) that will satisfy it. (Section \ref{ssec:dmdocs_review}) \\
\textbf{(4)} Find out what the expected filter change time will be (maximum is 120s). (Section \ref{ssec:dmdocs_review}) \\
\textbf{(5)} Confirm whether alternative products can contribute to the Alert Stream. (Section \ref{ssec:issues_alerts}) \\
\textbf{(6)} Update the database schema for {\tt DIASource} to add an element that contains information about which template was used to create the difference image (e.g., {\tt tractId} or {\tt patchId}). (Section \ref{ssec:issues_databases}) \\
\textbf{(7)} Clarify whether database schema key {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry. (Section \ref{ssec:issues_databases}) \\
\textbf{(8)} Find out whether a SP visit with a single exposure (no snaps) disqualifies it from incorporation into Level 1. (Section \ref{ssec:issues_snaps}) \\
\textbf{(9)} Confirm that the ISR algorithms will cover a range of exposure times and dither pattersn. (Section \ref{ssec:issues_calibrations}) \\
\textbf{(10)} Implanting fake sources remains TBD. (Section \ref{ssec:issues_calibrations}) \\
\textbf{(11)} Confirm that science users could write a Level 3 SAS with the existing planned algorithms? (Section \ref{ssec:issues_mops})

\noindent \textbf{Science Collaborations: } \\
\textbf{(1)} SC should be solicited to provide input about all special program science goals that require 60 second alert latency from non-standard exposure times. (Section \ref{ssec:issues_alerts})\\
\textbf{(2)} SC should be solicited to provide input on whether the space allotted for object characterization parameters in the database schema such as variability and morphology are adequate to describe objects with observations over a wider range of cadences and depths than the WFD observations. E.g., can the variability parameter characterize fast cadence data, does the morphology parameter appropriately identify low-surface brightness features. DM will first compile and provide a list of planned parameters from the schema. (Section \ref{ssec:issues_databases}) \\
\textbf{(3)} SC should be solicited to provide input on whether there are science use-cases for specialized, auto-generated template images e.g., with constant $\Delta t$, or image-image differencing (instead of image-template). DM should then assess whether this will require additional database elements. (Section \ref{ssec:issues_templates}) \\

\subsection{Interacting with the Science Community} \label{ssec:intro_forum}

Several of the Action Items suggest soliciting input from the science collaborations (SC). However, it may be wiser/easier to wait until the next call for community input, and package these requests up in a broader request that a kind of "DM Requirements" section be included in all white papers. The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.

@@

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\clearpage
\section{The Current DM Plans for Special Programs} \label{sec:dmdocs}

In this section we review what is written down in LSST documents with regards to the processing of Special Programs data. Below, we review the relevant contents of the three Data Management documents, and then the non-DM LSST documents.

\textbf{Overview:} Although the data from some special programs might be incorporated into the Level 1, 2, and/or MOPS pipelines (see Section \ref{ssec:dmdocs_SPinWFD} below), every special program will also have one or more associated Level 3 processing pipelines. These Level 3 processing pipelines will be assembled by the science users, hopefully using existing LSST codes and schemas (so that e.g., the Level 3 databases that can be joined to the Level 1/2 databases), and also with custom codes. It is clearly specified that DM will not write any specialized algorithms for Special Programs data, but that an effort will be made to ensure the DM codes are extendable to Level 3. DM will also ensure that the relevant software infrastructure exists for science users, and for LSST to automatically run user-provided Level 3 pipelines (e.g., difference imaging on deep drilling nightly stacks). To enable science goals, some Level 3 pipelines can be run by LSST with a projected latency of (at most) tens of minutes (as opposed to the 60 second guaranteed for WFD). Level 3 pipelines for stacks and co-adds that do not have the same real-time urgency can simply be run by users through the Science Platform.

\subsection{Documentation Review}\label{ssec:dmdocs_review}

\noindent \textbf{DMSR, LSE-61} \\
Data Management Subsystems Requirements \citep{LSE-61} contains the high-level requirements for all processing and products. These have flowed down from the SRD (LPM-17), the SR (LSE-29), and the OSS (LSE-30), which are discussed in the next section. \\
$\bullet$ The only part of the DMSR that directly references Special Programs (or DDF or mini-surveys) is Section 1.2.2 "Processed Visit Images" (ID: DMS-REQ-0069), which specifies that "Processed science exposures are not archived, and are retained for only a limited time to facilitate down-stream processing. They will be re-generated for users on-demand using the latest processing software and calibrations", but also says that "This aspect of the processing for Special Programs data is specific to each program". \\
$\rightarrow$ \textcolor{red}{It is unclear whether processed SP data will remain on disk and accessible when required by the science users, or if it will only live a short while}. \\
$\rightarrow$ \textbf{Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request.} \\
$\bullet$ Section 2.9, "Level 3 Production", specifies high-level services of the DMS for Level 3 processing, which are relevant to special programs. \\
2.9.1. Level 3 Data Import (ID: DMS-REQ-0290): to provide the ability to ingest common file formats (e.g., FITS table of external catalog). \\
2.9.2. Resource Allocation (ID: DMS-REQ-0119): to provide a mechanism to prioritize Level 3 activities and appropriately allocate processing resources. \\
2.9.3. Data Product Self-Consistency (ID: DMS-REQ-0120): to provide a means for ensuring Level 3 tasks can be carried out on self-consistent inputs. $\rightarrow$ \textcolor{red}{(?).} \\
2.9.4. Provenance (ID: DMS-REQ-0121): to provide a means for recording information about processing performed at DACs (DMS-provided input and user inputs) to help users work in a reproducible way. \\
2.9.5/6. Software Frameworks (ID: DMS-REQ-0125,DMS-REQ-0128): to provide a means for applying user-provided processing to catalog and image data, and for assessing the completeness of that application. \\
3.1. Software for Community Re-Use (ID: DMS-REQ-0308): algorithms shall be designed to run on both high-performance platforms and users' desktops. $\rightarrow$ \textcolor{red}{I thought that the Science Platform was being developed specifically so that no one needs to process LSST data on their home machines?} \\

\noindent \textbf{DMAD, LDM-151}\\
LSST Data Management Applications Design \citep{LDM-151} outlines the code and algorithms that will be used in the processing of images in the AP (Level 1), DRP (Level 2), and MOPS. As such, it does not say much about Special Programs. It does say that "LSST DM is required to facilitate the creation of Level 3 data products by providing suitable APIs, software components, and computing infrastructure, but will not by itself create any Level 3 data products." The latter meaning that DM will not create any processing algorithms that are specific to the needs of any Special Programs data. Section 7 briefly describes the Data Access and the Science User Interface and Tool (SUIT), also known as the Science Platform, with a mention that they will be described in full in later documents. The Data Access layer (DAX) will be for sharing, accessing, searching (etc.) products, and the SUIT/Platform will provide an interactive environment for users, and also serve as portal for Level 3 pipeline development (including installed version of the LSST Software Stack), execution, retrieval, and analysis.

\noindent \textbf{DPDD, LSE-163}\\
Data Products Definitions Document \citep{LSE-163} specifies that the data products for Special Programs will ``be created using the same software and hardware as Levels 1 and 2". Section 7 is devoted to the data products of Special Programs, explaining how the current plan is to have both the Level 1 and 2 processing pipelines run on SP data in the same way that they run on the WFD, and produce all of the same data products and alerts, when possible/desirable. The Level 3 products for SP data can/will be kept in separate (joinable) databases that have identical schema as the WFD databases. \\
$\bullet$ It is specified that SP data processing will be limited to not use more than 10\% of the DM computational and storage facilities. During the telecon on Wed Mar 29, KT Lam reviewed the computational budget but it seemed mainly for WFD with only passing mention of DDF or special programs. \\
$\rightarrow$ \textcolor{red}{10\% of the computational budget seems like an underestimate. Although SP data will not be more than 10\% of the total data volume, it will be multiply processed: once as WFD Level 1, again perhaps as Level 2, and then possibly in multiple pipelines as Level 3, some of which are extremely computationally intensive (e.g., shift-and-stack). On top of this, add the Level 3 processing that will be applied to WFD data, and we could easily find that the Special Programs data are processed $\sim4\times$ as much as WFD.} \\
$\rightarrow$ \textbf{Confirm if the budget requires additional computational resources for Special Programs and Level 3 processing.}

\noindent \textbf{SRD, LPM-17}\\
The Science Requirements Document \citep{LPM-17} makes no reference to special programs, deep drilling fields, or mini-surveys; it's scope is the science of the WFD only.

\noindent \textbf{LSST SR, LSE-29}\\
The System Requirements document \citep{LSE-29} is derived from the LSST Science Requirements Document (SRD, \citep{LPM-17}). It contains the high-level specifications about fraction of total observing time delegated for special programs, minimum exposure times, maximum filter change time, etc., that flow down to the DMSR and OSS (detailed and discussed above and below).

\noindent \textbf{OSS, LSE-30}\\
The Observatory System Specifications (OSS) document \citep{LSE-30} contains high-level requirements for the Level 3 data products, and their storage and access, that are similar to those in the DMSR (as they both stem from LPM-17 and LSE-29). There are two points worth mentioning: \\
$\bullet$ 3.6.1.4. Minimum exposure time (ID: OSS-REQ-0291): 1 second (stretch 0.1 seconds). \\
$\rightarrow$ \textcolor{red}{There is a note that if the exposure time is shortened from 15 seconds, the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that thermal stability might also be affected with longer exposure times.}\\
$\rightarrow$ \textbf{Find out the thermal tolerances and determine the minimum and maximum exposure times (and/or cadence of short exposures) that will satisfy it.} \\
$\bullet$ 3.6.2.2. Maximum time for operational filter change (ID: OSS-REQ-0293): 120 seconds \\
$\rightarrow$ \textcolor{red}{This maximum time for operational filter change would be a debilitating overhead for a deep drilling cadence of consecutive 30 second images that rotates through filters.}\\
$\rightarrow$ \textbf{Find out what the expected filter change time will be.} \\
$\bullet$ 3.6.2.4. Minimum filter change count (ID: OSS-REQ-0295): day, 8, night, 4 \\
$\rightarrow$ \textcolor{red}{There is no related specification limiting the maximum number of total filter changes to avoid wear-and-tear on the mechanism?}

\noindent \textbf{LSE-180}\\
Level 2 Photometric Calibration for the LSST Survey \cite{LSE-180} is built on OpSim runs that do include some nominal deep drilling fields, but the photometric calibration investigated in this work does not deal with potential issues induced by non-standard visit patterns or exposure times of special programs -- it's scope is that of the WFD Main Survey.

@@

I haven't found much written down in terms of guidelines or expectations for incorporating Special Programs data into the Level 1 and Alert Stream, Level 2 DRP, and/or MOPS pipelines -- especially when these data deviate from the standard 30 second visit -- but it seems feasible. Here we make these considerations.

\noindent \textbf{Incorporating SP data into Level 1 and the Alert Stream.} \\
It is reasonable to attempt to include as many LSST images into Level 1 and the Alert Stream as possible. When the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, this is obviously fine. When the exposure time is shorter than a WFD visit image, this should be fine as well. The minimum supported exposure time is currently specified as 1 second, which spans a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes. A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (all other filters have a similarly large overlap). The exposure times of a special program image might be longer than the WFD visit image, but given that the template will typically be a stack of a year or two of data (at least), and that it's hard to imagine a reason for even doing $150$ second exposures (given that the saturation would be $r\approx 18.3$ and the cosmic rays would intensify), it seems unlikely that any images would be rejected from Level 1 for having too long an exposure time. This means that the limiting factor of including any/all special programs data into the Level 1 and Alert Stream will be a processing back-up caused by the higher cadence of short exposures, or a lack of template images for special programs that go outside the WFD search area.

\noindent \textbf{Incorporating SP data into MOPS.} \\
Needs and capabilities are currently undefined.

\noindent \textbf{Incorporating SP data into the Level 2 DRP.} \\
The core science needs for uniform-depth images and catalogs for the WFD main survey probably means this will not happen, but identical Level 3 data pipelines and products may be used for regions with deeper coverage and the two databases can be joined.

>>>>

\documentclass[DM,lsstdraft,toc]{lsstdoc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{latexsym}
\usepackage{color}
% black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, blue, orange, pink, purple, red, teal, violet, white, yellow.
\usepackage{enumitem}

@@

\author{M.~L.~Graham and M.~Juri\'{c}}

@@

\setDocAbstract{\textbf{WORKING DRAFT (contains rough notes and partial thoughts) } but intended to be an overview of any and all potential requirements on data management (DM) from Special Programs such as the deep drilling fields (DDF) or mini-surveys. We also summarize the plans for how the Special Programs data will be integrated into the wide-fast-deep (WFD) Main Survey (Levels 1 and 2). It is the intent that this document evolve to become an internal catalog of change requests to the DM Level 1 and 2 pipelines and products, and a community resource for future Special Program white papers and preparations for Level 3 pipelines. }

@@

\noindent The purpose of this study is to: \\
(1) summarize the current plans of the DM pipelines with respect to Special Programs, \\
(2) summarize the processing that is required to enable science for Special Programs, \\
(3) assess whether the DM pipelines meet the expected needs of Special Programs, and \\
(4) ensure any necessary changes get written into the requirements, designs, and plans.

\noindent This will be accomplished by: \\
(1) starting with the DPDD, whitepapers, etc., and compiling relevant information, \\
(2) discussing initial issues internally with DM (Mario, the SST), engineering (K-T, Tim), and then \\
(3) ensuring the identified needs become specified requirements (i.e., spawn change requests).

@@

% % % % % % % % % % % % % % % % % % 
\subsection{Boundaries of Allowed Special Programs Observations}\label{ssec:intro_bounds}

In the next couple of years, cadences and pointings for DDF and new mini-surveys will be proposed by the community. The LSST project will need to define boundaries for these observations. Although not all of these boundaries will originate with DM constraints, we start that process here (\textcolor{red}{MLG: Are any initially derived boundaries for Special Programs observations specified elsewhere?}). The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3}

Start of a list of observing restrictions. \\
\textbf{Sky coordinates --} (The RA and Dec that will be accessible from the site.) \\
\textbf{Pointing alt/az, airmass --} (Telescope pointing constraints.) \\
\textbf{Sky background level --} (e.g., twilight observations cannot start before X) \\
\textbf{Exposure times --} (e.g., current shortest is 1 s, with stretch goal of 0.1, but... \\
\textbf{Filters --} (e.g., list filter combinations; $gri$ always be on?; change times, etc.) \\
\textbf{Afternoon calibrations --} (e.g., must fit into some amount of time left over after the WFD main survey daily calibration frames are obtained) \\
\textbf{Time allotted --} Nominally, 10\% of the total observing time (90\% goes to the WFD Main Survey). \\

% % % % % % % % % % % % % % % % % % 
\subsection{Interacting with the Science Community} \label{ssec:intro_forum}

To resolve some open issues we will need to solicit input from the science collaborations (SC). However, it may be better and easier to wait until the next call for community input, and package these requests up in a broader request that a kind of "DM Requirements" section be included in all white papers. The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.

@@

% % % % % % % % % % % % % % % % % % 
\subsection{A Summary of Open Action Items} \label{ssec:intro_AI}

\begin{enumerate}[topsep=-15pt,label= \textbf{Action \Roman*.}] \item Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications. \end{enumerate}
\begin{enumerate}[topsep=-15pt,label= \textbf{Action \Roman*.},resume] \item Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Query the science community and NCSA about the science needs and processing capabilities, respectively, of $<30$ second cadence imaging. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on whether {\tt float[6 x 32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on the characterization parameters for extended objects. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the lower and upper limits on exposure time. \end{enumerate}

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\clearpage
\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmdocs}

In this section we summarize how DM intends to process the data obtained during Special Programs observations. This high-level "intent" is a combination of specifications that are actually written down in DM-related documents, and unwritten plans that exist in e.g., Mario's Brain. {\it MLG: Note that Sections \ref{ssec:dmdocs_SPinWFD} and \ref{ssec:dmdocs_L3} currently contain a mixture of what is officially written down and what is unofficially planned or expected, usually my interpretation of a conversation with Mario, while Section \ref{ssec:dmdocs_review} is a direct review of the DM documentation. Once the SST and engineering have converged on a treatment plan for Special Programs data via this work, my plan is to return to the official DM documentation and identify where change requests need to be initiated.}

\textbf{Overview:} Although the data from some Special Programs might be incorporated into the Level 1, 2, and/or MOPS pipelines, every Special Program will probably also need one or more associated Level 3 processing pipelines. These Level 3 processing pipelines will be assembled by the science users, hopefully using existing LSST codes and schemas so that e.g., the Level 3 databases that can be easily joined to the Level 1 and 2 databases. It is clearly specified that DM will not write any specialized algorithms for Special Programs data, but that DM will ensure all pipeline code is extendable to Level 3 and that the relevant software infrastructure exists for science users to create and run their Level 3 pipelines. In addition, LSST will "federate" some Level 3 pipelines, putting them under change control and taking responsibility for their processing (e.g., shift-and-stack for faint moving objects is expected to become a federated Level 3 processing). All user access to processing codes and data products will be through the Science Platform.

\textbf{Section contents}: \\
Section \ref{ssec:dmdocs_SPinWFD} -- an overview of whether and how DM will incorporate Special Programs data into the Level 1, 2, and MOPS pipelines and products. \\
Section \ref{ssec:dmdocs_L3} -- an overview of how federated and user-run Level 3 pipelines will process Special Programs data, incorporate WFD data, and/or join to Level 1 and 2 products. \\
Section \ref{ssec:dmdocs_review} -- a detailed break-down of Special Programs (or Level 3) related content in the three DM documents (DPDD, DMSR, and DMAD), and non-DM LSST documents. \\

@@

There is not much, if anything, specified regarding guidelines or expectations for incorporating Special Programs data into the Level 1 and Alert Stream, Level 2 DRP, and/or MOPS pipelines that will automatically run on all WFD main survey observations. The concern is that drastic changes in the exposure time or total depth of the observations might negatively impact the science goals that depend on the WFD main survey data products. Here we summarize these concerns.

\subsubsection{Level 1 and the Alert Stream}\label{ssec:dmdocs_SPinWFD_L1}

It is beneficial to transient science to include as many LSST images into Level 1 and the Alert Stream as possible. Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream (e.g., no Level 3 difference imaging pipeline can send its detections to the Alert Stream). Only images that qualify for Level 1 processing will contribute to the Alert Stream. The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a SP image's suitability for Level 1 is to be in a region of sky with an existing template. So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is obviously going to be fine. However, the suitability of shorter and longer exposures is unknown. Here we consider the suitability -- or what we need to do in order to determine the suitability -- of shorter and longer exposures for the Level 1 and Alert Stream. These considerations are made independently of the science motivations of the Special Programs that proposed the different exposure times.

% Shorter exposures: twilight survey, getting bright stars on the same photometric system, variability in brighter stars
% Longer exposures: the weak lensing community might propose long-exposure fields if they remain concerned that 2x15 seconds is insufficient to obtain a stable PSF with no spatial variation dominated by the atmosphere

(1) Shorter Exposures. The minimum supported exposure time is currently specified as 1 second (stretch goal 0.1 seconds), but this is a camera constraint. The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint, but is currently unspecified. We consider this issue separately in Section \ref{ssec:issues_calibrations}. Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes. A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap). However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale).  Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.

(2) Longer Exposures. There is no maximum exposure time specified for an LSST image. Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine. However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources. With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).

In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream. One possibly useful study is Chang et al. (2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data". They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize. They present a new software package to do mitigate the effects. Alternatively, we may need to use software packages {\tt PhoSim} (Peterson et al. 2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or {\tt ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.

\begin{enumerate}[topsep=-10pt,label= \textbf{Action \Roman*.}] \item Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream. \end{enumerate}

\subsubsection{MOPS}\label{ssec:dmdocs_SPinWFD_MOPS}

The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data. While it is expected that MOPS will be able to identify appropriate images and incorporate as much Special Programs data as possible, the capabilities are currently undefined, and we leave it for the MOPS pipeline builders to decide. We note that since MOPS does not run during the night, it does not have the same latency requirements, or the associated issues.

\subsubsection{Level 2 Data Release Pipeline}\label{ssec:dmdocs_SPinWFD_L2}

Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible. We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP {\tt Objects} database that is built from a set of images with nearly constant depth and cadence. When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP. For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions. In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs.

% % % % % % % % % % % % % % % % % % 
\subsection{SP Data Processing with Federated and User-Run Level 3}\label{ssec:dmdocs_L3}

Through the Science Platform, member of the science community can assemble Level 3 processing packages using both LSST and custom codes. The DM team is required to write their codes to be extendable to Level 3, and to provide the infrastructure for users to run them on LSST servers, monitor their progress, and analyze their output (including a prioritization system to allocate processing resources). All Level 3 pipelines will have access to processed single visits from both the WFD main survey and Special Programs, as well as difference images, CoAdds, and catalogs from the Level 1 and 2 pipelines. It is expected that some Level 3 pipelines will become "federated", installed and operated (and change controlled) by the LSST Operations team, such as those developed for time domain science that need to run automatically and repeatedly. For both federated and user-run code, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of Level 1 and 2.

Level 3 pipelines for Special Programs data for time-domain astronomy will need access to images on a timescale from minutes to weeks after acquisition, as opposed to static-sky Level 3 pipelines that may run on e.g., a yearly schedule like the Level 2 DRP. It is currently unclear what that real-time latency will be, and how processing resources will be allocated; here we break down some of the options.

If the images obtained with a Special Program are incorporated into Level 1 (see Section \ref{ssec:dmdocs_SPinWFD}), then the Level 3 processing can begin as soon as the Level 1 imaging products are available: raw images, processed single visits (a.k.a. calexps), difference images, and their source catalogs. The current maximum time for this is {\tt \textbf{L1PublicT}}, specified by LSE-29 (\citep{LSE-29}) to be 24 hours (LSR-REQ-0104) -- but this seems far too long, especially for federated Level 3 pipelines, given that they are finished within $60$ seconds.

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs. \end{enumerate}

If the images obtained with a Special Program cannot be incorporated into Level 1, then the latency with which they will be made available is unspecified. Ultimately, their delivery will depend why they are not incorporated into Level 1, which will probably fall into two categories: (1) there is no template at their location, or (2) their exposure time prohibits difference imaging. In the former case, it seems likely that single processed images can be provided on a timescale of minutes, but this needs to be specified. In the latter case, it will depend on the exposure time. As mentioned in Section\label{ssec:dmdocs_SPinWFD_L1}, the minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint, but is currently unspecified, and discussed separately in Section \ref{ssec:issues_calibrations}. It may be that single-visit processing is simply not supported by DM below some exposure time threshold, and so the latency on raw image delivery should be specified. Although it is likely that slightly shorter/longer exposure images can be reduced and calibrated with the same pipeline as $15$ second exposures, even if they are unsuitable for difference imaging, this has not yet been confirmed, and needs to be investigated and specified.

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline. \end{enumerate}

In the DPDD \citep{LSE-163}, it specifies that Special Programs data processing will be limited to not more than 10\% of the DM computational and storage facilities. During the telecon on Wed Mar 29, KT Lam reviewed the computational budget, but it seemed mainly for WFD with only passing mention of DDF or Special Programs. This 10\% of the computational budget seems like an underestimate: although SP data will not be more than 10\% of the total data volume, it will be multiply processed: once as WFD Level 1, again perhaps as Level 2, and then possibly in multiple pipelines as Level 3, some of which are extremely computationally intensive (e.g., shift-and-stack). It seems that we could easily find that Special Programs data are processed $\sim4\times$ as much as WFD. Furthermore, it is unclear whether the location for all this processing has been identified; Mario asks, for federated real-time Level 3 processing, {\it "whether this the same batch system we make available to the users for running Level 3 codes or the one that?s used to process calibrations. Have these systems been sized?"}.

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place. \end{enumerate}

\subsection{In-Depth Documentation Review}\label{ssec:dmdocs_review}

In this section we attempt to track the flow-down requirements through the official LSST and DM documentation. We find some hardware-related constraints in the OSS that will set the maximum possible divergence of Special Programs raw images from WFD main survey images (i.e., exposure times, rate of filter changes), and these will be relevant to DM processing. We also find that some broad, high-level requirements for Level 3 and Special Programs are written down in the DMSR. The next document down-flow, the DMAD, is focused on the Level 1 and 2 algorithms, and although its Section 7 mentions the DAX and SUIT, it says that they are to be detailed in a different document. The DPDD also does not have much content that directly pertains to Special Programs and Level 3 processing.

\noindent \textbf{Science Requirements Document, SRD, LPM-17, \cite{LPM-17}}\\
The SRD makes no direct reference to Special Programs, deep drilling fields, or mini-surveys. It's scope is the science of the WFD only.

\noindent \textbf{LSST System Requirements, SR, LSE-29, \cite{LSE-29}}\\
The SR is derived from the SRD. It contains the high-level specifications about fraction of total observing time delegated for Special Programs, minimum exposure times, maximum filter change time, etc., that flow-down to and are repeated in the DMSR and OSS (discussed below).
